### 2022-04-04 manuscript.tex
april4 [ci skip]
This build is based on
https://github.com/scherzerthesis/thesis/commit/86f7aa7d5412b0559c0590dc4ab93fef7fb616ae.

This commit was created by the following CI build and job:
https://github.com/scherzerthesis/thesis/commit/86f7aa7d5412b0559c0590dc4ab93fef7fb616ae/checks
https://github.com/scherzerthesis/thesis/actions/runs/2091271672

This manuscript
(\href{https://scherzerthesis.github.io/thesis/v/14cf100164c7bfd47ddf716d89a664eaec87e10e/}{permalink})
was automatically generated
from \href{https://github.com/scherzerthesis/thesis/tree/14cf100164c7bfd47ddf716d89a664eaec87e10e}{scherzerthesis/thesis@14cf100}
on April 2, 2022.

@@

@@

@@

\#\#Abstract

@@

We also failed to detect robust differences between genotypes that are known to cause malignant transformation, such as p53 loss, p53 missense mutation, CDKN2A loss, or mutant PI3K\textsuperscript{H1047R} (Dankort et al., 2007; Trejo et al., 2013; van Veen et al., 2019). This is likely a result from a combination of factors, such as time of collection and library capture/sequencing depth. One straightforward hypothesis is that we did not allow for full transcriptomic changes to take effect that is imparted by tumor suppressor or oncogene mutation. Therefore, a timecourse analysis that spans early timepoints (4-6 weeks) and late timepoints (\textgreater16 weeks) will likely reveal more informative transcriptional heterogeneity. Indeed, our previous studies have indicated that there is no tumor growth advantage at early time points as late as 6 weeks post tumor initiation. One other possibility that could explain our lack of genotype-specific gene expression programs is that we did not capture enough unique transcripts in order to reveal the full extent of tumor heterogeneity. This is likely because there are known limitations to 10X genomic library preperation. Specifically, there is gene-dropout effects where only the most highly expressed genes are captured by the droplet-based technique. Indeed, 10X genomics technologies have been shown not to detect genes with lower expression(Wang et al., 2021). However, 10X genomics allows you to profile more cells than similiar scRNA-seq techniques. Genes with lower expression can be just as important as highly expressed genes for a cell's identity. Actually, transcription factors that determine cell states have been shown to have modest expression(Wang et al., 2021) Therefore, we hypothesis that using plate-based scRNA-seq approaches, such as Smart-Seq2, would allow for more gene capture. Thus, Smart-seq2 will enable us to more precisely determine transcirptional heterogenity and genotype-specific gene expression changes.

@@

\begin{figure}
\hypertarget{fig:04}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/scrna_3.png}
\caption{Tumor cells cells are clustered based on treatment}\label{fig:04}
}
\end{figure}

\begin{figure}
\hypertarget{fig:04}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/scrna_4.png}
\caption{P53 transcriptional signature is identified in BRAV\textsuperscript{V600E}, p53-wildtype tumor cells}\label{fig:04}
}
\end{figure}

\begin{figure}
\hypertarget{fig:04}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/scrna_5.png}
\caption{Differentially expressed genes imparted by mutant-p53}\label{fig:04}
}
\end{figure}

>>>>

This manuscript
(\href{https://scherzerthesis.github.io/thesis/v/86f7aa7d5412b0559c0590dc4ab93fef7fb616ae/}{permalink})
was automatically generated
from \href{https://github.com/scherzerthesis/thesis/tree/86f7aa7d5412b0559c0590dc4ab93fef7fb616ae}{scherzerthesis/thesis@86f7aa7}
on April 4, 2022.

@@

\textbf{Abstract}
Lung cancer remains a threat to human health throughout the world. As a conseqeunce of accumulation of mutations, from tobacco or otherwise, lung cancers have a vast mutational spectrum that affects many genes. These characteristics make lung cancer a challenge in understanding how the disease is formed. Furthermore, mutational tumor heterogeneity is linked to drug resistance and a poor prognosis for patients. Interestingly, MAPK pathway activation appears to be a prerequisate for the formation of lung adenocarcinoma. In chatper 1, using genetically engineered mouse models of BRAF\textsuperscript{V600E} lung tumorigenesis, we analyzed the contribution of commonnly mutated genes to tumor pregression. It was found that, while BRAF\textsuperscript{V600E} drives benign tumor formation, loss of SETD2 function drives tumor progression to larger tumors. Thus, we identified SETD2 as a common alteration in BRAF\textsuperscript{V600E} lung cancer that is a bone fide tumor supressor.

Genetic sequencing of human lung adenocarcinomas has also reavealed the widespread inactivation of p53 as a common alteration in late stage lung adenocarcinoma. Normal p53 function restricts tumor plasticity and thus losing p53 function enables tumor cells to adopt more malignant cell identities. However, not all p53 mutations are created equal. In chapter 2, we characterized the effect of three p53 mutant alleles (P53-null, P53-R172H, and p53-R245W) on BRAF\textsuperscript{V600E} lung tumorigenesis. It was found that p53-R172H and p53-R245W induce more widespread tumor dedifferentiation thus leading to worst tumor outcomes, compared to p53-loss. Cell lines derived from the tumors allowed for the dissection of signaling pathway dynamics which led us to observed higher MAPK pathway activation in p53-R172H and p53-R245W cell lines. We also found that p53-R172H and p53-R245W cell lines were more refractory to MAPK pathway inhibition with small molecule inhibitors. These altered phenotypes seen in p53-R172H and p53-R245W are liekly a reflection in their alter transcriptomic profiles.

Although genetically-engineered mouse models are foundational in understanding how lung adenocarcinomas delveop, progress, and metastasize, little is known about the transcriptional heterogeneity of genetically-engineered mice. In Chapter 3, we profiled the single-cell transcriptomes of BRAF\textsuperscript{V600E} lung tumors. We identified distinct clusters of tumor cell types and determined the transcriptional response to MAPK pathway-targeted therapy.

By quantifying the tumor suppresive landscape and the heterogenous transcriptional landscape of BRAF\textsuperscript{V600E} cells, our work has contributed valuable knowledge to the field of lung adenocarcinoma biology.

@@

Large-scale sequencing of cancers have revealed that lung adenocarcinomas have a high mutation rate. Although it is clear that the evolution of lung adenocarcinoma is centered around mutational activation of the EGFR-RAS-RAF-MEK-ERK pathway, little is known about secondary mutations that drive disease progression. Since lung cancers are riddled with mutations, it is difficult to glean which mutations support cancer growth and which are merely bystanders. Thefore, pre-clinical mouse models that are syngenetic are important models to determine oncogenicity of a particular gene. Here, we use Tuba-Sequencing combined with CRISPR/Cas9 gene editing to simmuletaneuosly determine the contribution of the loss of one of eleven tumor suppressor genes in BRAF\textsuperscript{V600E} mouse lung tumorigenesis. We show that loss of either \emph{Apc} or \emph{Setd2} leads to enhanced tumor growth. We validated SETD2 loss as a cooperating mutation with single-guide RNAs. FUnctionalily, SETD2 is a histone lysine methyltransferase that marks active gene bodies by depositing me\textsuperscript{3} marks on H3K36. In BRAF\textsuperscript{V600E} lung tumors, \emph{Setd2} loss decreases H3K36me\textsuperscript{3} and H3K36me\textsuperscript{3} was observed to correlate with larger tumors. \emph{SETD2} RNA is expressed less human lung cancers than normal lung and can be indicative of worse overall survival. Moreover, \emph{SETD2} loss is a common event in humna \emph{BRAF\textsuperscript{V600E}} lung adenocarcinomas. Together, our results highlight the potential importance of \emph{SETD2} in the progresion of BRAF\textsuperscript{V600E} lung adenocarinoma.

@@

\#\#Abstract
malignant lung adenocarcinomas tumors are initiated from a single neoplastic cell. Yet, there is substantial cellular heterogenity in a fully malignant tumor. Since human lung adenocarcinomas are heavily mutated, it is diffuclt to know wheter underlying cellular heterogenity is a consequence of the underlying mutational heterogenity. Mouse models of lung adenocarcinoma accurately represent genomic and histological aspects of human lung adenocarcinoma. However, genetically-engineered mouse model tumor heterogenity remains poorly understood. Here, we profile the single cell transcriptomes of genetically-engineered mouse models that are driven by the BRAF\textsuperscript{V600E} oncoprotein with 5 seperate cooperating genetic events ( \emph{Pi3k\textsuperscript{H1047R}}, \emph{CdkN2a\textsuperscript{fl/fl}}, \emph{Trp53\textsuperscript{fl/fl}}, \emph{Trp53\textsuperscript{R172H/R162H}}, and \emph{Trp53\textsuperscript{R245W/R245W}}). There are stereotypically distinct clusters of cells thorughout all genotypes of BRAF\textsuperscript{V600E}-driven tumors. We also identified known tumor cell identities in both human and mouse lung adenocarcinoma.

@@

We also failed to detect robust differences between genotypes that are known to cause malignant transformation, such as p53 loss, p53 missense mutation, CDKN2A loss, or mutant PI3K\textsuperscript{H1047R} (Dankort et al., 2007; Trejo et al., 2013; van Veen et al., 2019). This is likely a result from a combination of factors, such as time of collection and library capture/sequencing depth. One straightforward hypothesis is that we did not allow for full transcriptomic changes to take effect that is imparted by tumor suppressor or oncogene mutation. Therefore, a timecourse analysis that spans early timepoints (4-6 weeks) and late timepoints (\textgreater16 weeks) will likely reveal more informative transcriptional heterogeneity. Indeed, our previous studies have indicated that there is no tumor growth advantage at early time points as late as 6 weeks post tumor initiation. One other possibility that could explain our lack of genotype-specific gene expression programs is that we did not capture enough unique transcripts in order to reveal the full extent of tumor heterogeneity. This is likely because there are known limitations to 10X genomic library preperation. Specifically, there is gene-dropout effects where only the most highly expressed genes are captured by the droplet-based technique. Indeed, 10X genomics technologies have been shown not to detect genes with lower expression(Wang et al., 2021). However, 10X genomics allows you to profile more cells than similiar scRNA-seq techniques. Genes with lower expression can be just as important as highly expressed genes for a cell's identity. Actually, transcription factors that determine cell states have been shown to have modest expression{[}Wang et al. (2021); (\textbf{pmid?});30205475{]}.
Therefore, we hypothesis that using plate-based scRNA-seq approaches, such as Smart-Seq2, would allow for more gene capture. Thus, Smart-seq2 will enable us to more precisely determine transcirptional heterogenity and genotype-specific gene expression changes.

@@

\begin{figure}
\hypertarget{fig:04}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/scrna_3.png}
\caption{Murine lung adenocarcinoma transcripional programs in BRAF\textsuperscript{V600E} tumorigenesis}\label{fig:04}
}
\end{figure}

\begin{figure}
\hypertarget{fig:04}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/scrna_4.png}
\caption{Transcriptional heterogeneity of tumor cells}\label{fig:04}
}
\end{figure}

\begin{figure}
\hypertarget{fig:04}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/scrna_5.png}
\caption{P53 transcriptional signature is identified in BRAV\textsuperscript{V600E}, p53-wildtype tumor cells}\label{fig:04}
}
\end{figure}

\begin{figure}
\hypertarget{fig:04}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/scrna_5.png}
\caption{P53 transcriptional signature is identified in BRAV\textsuperscript{V600E}, p53-wildtype tumor cells}\label{fig:04}
}
\end{figure}

\begin{figure}
\hypertarget{fig:04}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/scrna_6.png}
\caption{Genotpye-specific gene expression changes.}\label{fig:04}
}
\end{figure}
